Locality and Computational Reliability of Linear Response Calculations for Molecular Systems

### 2018-11-26 AnalyticStructure_main.tex
added linguistic revision of the paper

Let us now inspect the case of LR calculations.
In this framework one consider the unperturbed Hamiltonian $\hnot$ and assume that
the GS density $\dmnot$ is accessible and expressed in a given computational treatment.
The linear response formalism allows us to evaluate the modification of the expectation value of a generic observable induced by a
time (or frequency)-dependent perturbing field $\op\Phi(\omega)$ acting on the system.
This quantity, written in the frequency domain, can be expressed through the evaluation of the \emph{linear response functional}
\be\lb{LinearResponseFunctDef1}
\braket{\delta\op O}_\Phi(\omega) = \trace{\dm'_\Phi(\omega)\op O} \;,
\ee
where $\op O$ represents the observable\footnote{For a sake of concreteness, we are limiting our considerations to observables that do not explicitly depend on time} under inspection
and we have introduced the \emph{response density operator} $\dm'_\Phi(\omega)$ that codifies the modification of $\dmnot$ induced by the perturbation $\op\Phi(\omega)$.

@@

Two distinct regimes can be identified, on the basis of value of $\omega$.
The first one, which we will refer to as the \emph{below threshold} regime, is realized when
$\omega$ is lower than the ionization potential $|\eps_h|$.
In this case all the FS are below their threshold level and equation \eqref{LinearResponseFunctDef2}
is expressed in terms of genuinely localized
quantities. At these frequencies, the linear response functional $\braket{\delta\op O}_\Phi(\omega)$
may thus be evaluated with computational setups that
are similar to the ones usually employed for GS calculations.
The convergence in this regime is ``strong'', in the sense that the basis functions employed enables us to express
with arbitrary precision \emph{both} the states $\ket{\op O \psi_p}$ and $\ket{f_p^\Phi(\omega)}$.
This is of course true even when the above states are only implicitly expressed by the employed LR treatment,
 as e.g. in the case of Krylov spaces generated from the states $\ket{\op\Phi(\omega) \psi_p}$ \cite{baroni2006,baroni2008,linlinKPM}.
 In this regime, a computational treatment based on localized basis set is susceptible to provide a precise answer, assuming a reasonable level of completeness.

@@

As an example in support of these arguments we compare the {dynamical polarizability $\alpha(\omega)$} of a $CO$ molecule computed in three distinct real-space computational setups, which differs for the dimension of the computational domain.
Results reported in the two panels of figure \ref{co_spectrum} show the real and imaginary part of $\alpha(\omega)$, respectively. In all the cases a typical behavior emerges: the curves are almost coincident in the below threshold regime,
whereas the effect of the computational setup are clearly visible for values of  $\omega$ that exceed the threshold value.
For these high-frequency regions, convergence might be reached for domains of much larger sizes than what presented here~\cite{baroni2008},
which are -- as already noticed -- more than sufficient below threshold.

@@

As a consequence, we deduce that a molecular excitation \emph{may} be expressed in terms of localized states
\emph{only} when its energy is lower than \emph{all} the ionization potentials of the occupied states participating to the excitation,
namely if $\Omega_a<|\eps_p|$ $\forall p \in a$. This will only be true if the operator
$\op V'[\op E_a]$ can be restricted to localized kets and bras.
If any of these hypothesis does not hold, the excitations of the systems are defined via genuinely delocalized wavefunctions.
Furthermore, excited states are subjected to the orthonormalization condition
\eqref{ExcitedStateOrthNormDef1} that acts as a ulterior constraint and has the effect of determining the \emph{quantization}
of the energy levels $\Omega_a$ associated to the bound-state-like objects.
For delocalized states, as in the case of the continuum states of $\hnot$,
such a condition has to be interpreted in distributional sense and does not lead to quantization of $\Omega_a$.
Thus we can conclude that all the states $\ket{\chi^a_p}$ and the $\ket{\phi^a_p}$ with $\Omega_a$ below the threshold value constitute
-- if localized -- a genuine \emph{discrete} set, while the ones in the unbound energy range behaves as generalized eigenvectors associated to a
continuum spectrum.

@@

So far we have showed with both formal arguments and numerical calculations that the excitation spectrum split into two sectors with very distinct features.
It is then reasonable to ask which is their relative importance and, in particular, if there is the possibility to express observable quantities, in a suitable $\omega$ range, by only employing the sector of genuinely localized excitations.
Stated otherwise, will the \emph{restriction} of $\sop \chi(\omega)$ to only localized excitations can be useful to extract some physical quantities?

@@

A given numerical treatment typically provides only an effective description of the sector of delocalized excitation which in turns give rise to an effective 
representation of the susceptibility functional. It appears therefore important to identify indicators that enable us to validate the capability
of $\sop \chi(\omega)$ to express the localised FS (and thus the complete response density $\dm'_\Phi(\omega)$) below the IP threshold.
%It appears therefore important to identify indicators that might assess the capability of
%$\sop \chi(\omega)$ to express localised FS (and thus the complete response density $\dm'_\Phi(\omega)$) below the IP threshold.
For instance, the static polarizability tensor $\alpha_{ij}$ can be calculated as
% \be\lb{sminus2}
% \alpha_{ij} = \sint \dd a \frac{w^a_{ij}}{\Omega_a^2}\,, \quad w^a_{ij} \equiv \sum_{p\in a} \bra{\psi_p} \op {\mathbf r}_i \ket{w_p^a}\bra{\psi_p} \op {\mathbf r}_j \ket{w_p^a}\;.
% \ee
\be\lb{sminus2}
\alpha_{ij} = \sint \dd a \frac{w^a_{ij}}{\Omega_a^2}\,, \quad w^a_{ij} \equiv \sum_{p\in a} |\bra{\psi_p} \op {\mathbf r}_i \ket{w_p^a}|^2 \;.
\ee
Such value can be straightforwardly compared with results of Eq.~\eqref{staticalpha}, where the localized FS is obtained from GS techniques as
described in Sec.~\ref{FluctuationState}.
This is the well-know $S_{-2}$ sum-rule (see e.g.\cite{Wagner2012}), that express the value of the static polarizability
in terms of the molecule's oscillator strengths.
In the context of a numerical evaluation of the excitations, such sum rule can therefore interpreted as the capability of the pseudo-continuum sector of the excitation to
express the localized FS -- and consequently the reference value of $\alpha_{ij}$ -- at $\omega=0$.

We have analized the fullfillment of Eq.~\eqref{sminus2} for moeculs of various
sizes and symmetries.
In Fig.~\ref{co_AlphaExc} we show explicitly such a comparison in the case of the $CO$ molecule.
We notice that a considerable number of pseudo-continuum excitations is required
to find the correct result; the sector of the discrete excitations, even though
parametrised by localized, observable (in the sense of reproducible) excited states,
appears to be largely insufficient to express the reference fluctuation state. Its contribution is even negligible w.r.t.
the one coming from the continuum sector.

@@

Such a scenario can be also found in the other molecules we have considered;
varying the molecule size and symmetry seems not to modify the qualitative behavior of the excitation landscape. In addition the
localized sector of the excitations alone is largely insufficient to express $\alpha_{ij}$.
Such considerations are independent on the number of atoms of the molecule, as the range in $\omega$ needed to
achieve satisfactory agreement is largely above the last valence ionization threshold, which is
a intensive quantity.

@@

We show that Casida equations are equivalent to the equations of motion for the excited states \eqref{ExcitationOperatorsDef3}. To this aim we introduce an explicit basis $\{\ket{s}\}$
in the subspace of empty states so that both $\ket{\phi^a_p}$ and $\bra{\chi^a_p}$ can be represented as
\be
\ket{\phi^a_p} = \sum_s X^a_{p s}\ket{s} \;, \qq
\bra{\chi^a_p} = \sum_s\bra{s}Y^a_{p s} \;. \lb{phiandchi}
\ee
Plugging this expansion into equations \eqref{ExcitationOperatorsDef3} and projecting on a arbitrary element of the basis provides a linear systems of equations for the coefficients
$X^a_{ps}$ and $Y^a_{ps}$, namely
\begin{align}
 \sum_{s'}(H_{0ss'}-\eps_p\delta_{ss'})X^a_{ps'} + \bra{s}\op V'[\op E_a]\ket{\psi_p} &= \Omega_a X^a_{ps} \;, \nn \\
 -\sum_{s'} (H_{0 s' s}-\eps_p\delta_{s s'})Y^a_{ps'} - \bra{\psi_p}\op V'[\op E_a]\ket{s}& = \Omega_a Y^a_{ps} \;. \nn
\end{align}
Using this basis, excitation operators are expressed as a linear combination of transition operators $\excite{p}{s} \equiv \ketbra{s}{\psi_p}$ and $\decay{s}{p} \equiv \ketbra{\psi_p}{s}$,
as follows
\be\lb{ExcitationOpBasisTransition1}
\op E_a = \sum_{ps}\excite{p}{s}X^a_{ps}+\decay{s}{p}Y^a_{ps} \;.
\ee
The contribution of the coupling operator in the equations of motion of excited states can expressed by means of the coupling kernel \eqref{CouplingKernelDef1} as
\begin{align}
& \bra{s}\op V'[\op E_a]\ket{\psi_p} = \sum_{qs'}\left(U[\decay{s}{p};\excite{q}{s'}]X^a_{qs'} + U[\decay{s}{p};\decay{s'}{q}]Y^a_{qs'}\right) \;,\nn \\
& \bra{\psi_p}\op V'[\op E_a]\ket{s} = \sum_{qs'}\left(U[\excite{p}{s};\excite{q}{s'}]X^a_{qs'} + U[\excite{p}{s};\decay{s'}{q}]Y^a_{qs'}\right) \;, \nn
\end{align}
Give that we employ real functions both for $p$ and $s$ is real we also have $\excite{p}{s}(\r,\r') = \decay{s}{p}(\r',\r)$ and the density operator is symmetric in $\r \leftrightarrow \r'$.
These conditions applied together imply that $U[\excite{p}{s};\excite{q}{s'}] = U[\decay{s}{p};\decay{s'}{q}]$, as well as $U[\decay{s}{p};\excite{q}{s'}] = U[\excite{p}{s};\decay{s'}{q}]$,
and so equations \eqref{ExcitationOperatorsDef3} can be recasted in matrix form
\be\lb{ExcitationMatrixEq1}
\sum_{qs'}\mat{F_{ps}^{qs'} &  D_{ps}^{qs'}  \\
- D_{ps}^{qs'} & - F_{ps}^{qs'} }
\mat{X^a_{qs'} \\ Y^a_{qs'}} = \Omega_a \mat{X^a_{ps} \\ Y^a_{ps}}
\ee
where
\begin{align}
& F_{ps}^{qs'} = (H_{0 ss'}-\eps_p\delta_{ss'})\delta_{pq} + U[\decay{s}{p};\excite{q}{s'}] \;,\nn \\
& D_{ps}^{qs'} = U[\decay{s}{p};\decay{s'}{q}] \;.
\end{align}
This is the well-known Casida Equation in the basis of transitions for the excitation of energy $\Omega_a$. The computational reliability of the
eigenvalue $\Omega_a$ depends on the capability of the basis set $\ket{s}$ to fulfill of Eq.~\eqref{phiandchi}. Note that this may be valid only for a subset of the solutions of the eigenproblem \eqref{ExcitationMatrixEq1}.

\section{Computational details} \lb{compdetails}
The illustrative calculations presented in this paper are all KS-DFT calculations at the LDA level of theory. We employed the BigDFT code \cite{BigDFT},
that make usage of Daubechies Wavelets as computational basis set. This orthonormal basis presents optimal features of in view of reproducibility of the results, and it has proven to represent precisely and explicitly system with Free as well as Periodic boundary conditions (BC), without any basis set superposition error nor supercell aliasing, for Free BC.

For Ground State results, the important parameters in the context of convergence of the calculations are the spacing of the wavelet grid and the size of the simulation domain: for bound-state like functions, convergence is achieved by reducing the grid spacing and increasing the simulation box size.
Norm Conserving Pseudopotentials (PSP) of the HGH \cite{NLCCPSP} type are employed to remove the core electrons. Such PSP have proven to be able to provide all-electron accuracy for molecular systems.

@@

\subsection{Benzene}
The molecule is oriented in the $xy$ plain and the $x$-dimension of the box is equal to 15.0, 18.2 and 19.7 \AA.
220 states have been computed in each setup and the corresponding
energy reange for the unoccupied states is 16.9, 11.6 and 9.7 eV.

>>>>

Let us now inspect the case of LR calculations.
In this framework one considers the unperturbed Hamiltonian $\hnot$ and assumes that
the GS density $\dmnot$ is accessible and expressed in a given computational treatment.
The linear response formalism allows us to evaluate the modification of the expectation value of a generic observable induced by a
time (or frequency)-dependent perturbing field $\op\Phi(\omega)$ acting on the system.
This quantity, written in the frequency domain, can be expressed through the evaluation of the \emph{linear response functional}
\be\lb{LinearResponseFunctDef1}
\braket{\delta\op O}_\Phi(\omega) = \trace{\dm'_\Phi(\omega)\op O} \;,
\ee
where $\op O$ represents the observable\footnote{For a sake of concreteness, we are limiting our considerations to observables that do not explicitly depend on time} under inspection
and we have introduced the \emph{response density operator} $\dm'_\Phi(\omega)$ that codifies the modification of $\dmnot$ induced by the perturbation $\op\Phi(\omega)$.

@@

Two distinct regimes can be identified, on the basis of value of $\omega$.
The first one, which we will refer to as the \emph{below threshold} regime, is realized when
$\omega$ is lower than the ionization potential $|\eps_h|$.
In this case all the FS are below their threshold level and equation \eqref{LinearResponseFunctDef2}
is expressed in terms of genuinely localized
quantities. At these frequencies, the linear response functional $\braket{\delta\op O}_\Phi(\omega)$
may thus be evaluated with computational setups that
are similar to the ones usually employed for GS calculations.
The convergence in this regime is ``strong'', in the sense that the basis functions employed enable us to express
with arbitrary precision \emph{both} the states $\ket{\op O \psi_p}$ and $\ket{f_p^\Phi(\omega)}$.
This is of course true even when the above states are only implicitly expressed by the employed LR treatment,
 as e.g. in the case of Krylov spaces generated from the states $\ket{\op\Phi(\omega) \psi_p}$ \cite{baroni2006,baroni2008,linlinKPM}.
 In this regime, a computational treatment based on localized basis set is susceptible to provide a precise answer, assuming a reasonable level of completeness.

@@

As an example in support of these arguments we compare the {dynamical polarizability $\alpha(\omega)$} of a $CO$ molecule computed in three distinct real-space computational setups, which differs for the dimension of the computational domain.
Results reported in the two panels of figure \ref{co_spectrum} show the real and imaginary part of $\alpha(\omega)$, respectively. In all the cases a typical behavior emerges: the curves are almost coincident in the below threshold regime,
whereas the effects of the computational setup are clearly visible for values of  $\omega$ that exceed the threshold value.
For these high-frequency regions, convergence might be reached for domains of much larger sizes than what presented here~\cite{baroni2008},
which are -- as already noticed -- more than sufficient below threshold.

@@

As a consequence, we deduce that a molecular excitation \emph{may} be expressed in terms of localized states
\emph{only} when its energy is lower than \emph{all} the ionization potentials of the occupied states participating to the excitation,
namely if $\Omega_a<|\eps_p|$ $\forall p \in a$. This will only be true if the operator
$\op V'[\op E_a]$ can be restricted to localized kets and bras.
If any of these hypothesis does not hold, the excitations of the systems are defined via genuinely delocalized wavefunctions.
Furthermore, excited states are subjected to the orthonormalization condition
\eqref{ExcitedStateOrthNormDef1} that acts as a ulterior constraint and has the effect of determining the \emph{quantization}
of the energy levels $\Omega_a$ associated to the bound-state-like objects.
For delocalized states, as in the case of the continuum states of $\hnot$,
such a condition has to be interpreted in distributional sense and does not lead to quantization of $\Omega_a$.
Thus we can conclude that all the states $\ket{\chi^a_p}$ and the $\ket{\phi^a_p}$ with $\Omega_a$ below the threshold value constitute
-- if localized -- a genuine \emph{discrete} set, while the ones in the unbound energy range behave as generalized eigenvectors associated to a
continuum spectrum.

@@

So far we have showed with both formal arguments and numerical calculations that the excitation spectrum splits into two sectors with very distinct features.
It is then reasonable to ask which is their relative importance and, in particular, if there is the possibility to express observable quantities, in a suitable $\omega$ range, by only employing the sector of genuinely localized excitations.
Stated otherwise, can the \emph{restriction} of $\sop \chi(\omega)$ to only localized excitations be useful to extract some physical quantities?

@@

A given numerical treatment typically provides only an effective description of the sector of delocalized excitation which in turns gives rise to an effective 
representation of the susceptibility functional. It appears therefore important to identify indicators that enable us to validate the capability
of $\sop \chi(\omega)$ to express the localised FS (and thus the complete response density $\dm'_\Phi(\omega)$) below the IP threshold.
%It appears therefore important to identify indicators that might assess the capability of
%$\sop \chi(\omega)$ to express localised FS (and thus the complete response density $\dm'_\Phi(\omega)$) below the IP threshold.
For instance, the static polarizability tensor $\alpha_{ij}$ can be calculated as
% \be\lb{sminus2}
% \alpha_{ij} = \sint \dd a \frac{w^a_{ij}}{\Omega_a^2}\,, \quad w^a_{ij} \equiv \sum_{p\in a} \bra{\psi_p} \op {\mathbf r}_i \ket{w_p^a}\bra{\psi_p} \op {\mathbf r}_j \ket{w_p^a}\;.
% \ee
\be\lb{sminus2}
\alpha_{ij} = \sint \dd a \frac{w^a_{ij}}{\Omega_a^2}\,, \quad w^a_{ij} \equiv \sum_{p\in a} |\bra{\psi_p} \op {\mathbf r}_i \ket{w_p^a}|^2 \;.
\ee
Such value can be straightforwardly compared with results of Eq.~\eqref{staticalpha}, where the localized FS is obtained from GS techniques as
described in Sec.~\ref{FluctuationState}.
This is the well-know $S_{-2}$ sum-rule (see e.g.\cite{Wagner2012}), that expresses the value of the static polarizability
in terms of the molecule's oscillator strengths.
In the context of a numerical evaluation of the excitations, such sum rule can therefore be interpreted as the capability of the pseudo-continuum sector of the excitation to
express the localized FS -- and consequently the reference value of $\alpha_{ij}$ -- at $\omega=0$.

We have analized the fullfillment of Eq.~\eqref{sminus2} for molecules of various
sizes and symmetries.
In Fig.~\ref{co_AlphaExc} we show explicitly such a comparison in the case of the $CO$ molecule.
We notice that a considerable number of pseudo-continuum excitations is required
to find the correct result; the sector of the discrete excitations, even though
parametrised by localized, observable (in the sense of reproducible) excited states,
appears to be largely insufficient to express the reference fluctuation state. Its contribution is even negligible w.r.t.
the one coming from the continuum sector.

@@

Such a scenario can be also found in the other molecules we have considered;
varying the molecule size and symmetry seems not to modify the qualitative behavior of the excitation landscape. In addition the
localized sector of the excitations alone is largely insufficient to express $\alpha_{ij}$.
Such considerations are independent on the number of atoms of the molecule, as the range in $\omega$ needed to
achieve satisfactory agreement is largely above the last valence ionization threshold, which is
an intensive quantity.

@@

We show that Casida equations are equivalent to the equations of motion for the excited states \eqref{ExcitationOperatorsDef3}. To this aim we introduce an explicit basis $\{\ket{s}\}$
in the subspace of empty states so that both $\ket{\phi^a_p}$ and $\bra{\chi^a_p}$ can be represented as
\be
\ket{\phi^a_p} = \sum_s X^a_{p s}\ket{s} \;, \qq
\bra{\chi^a_p} = \sum_s\bra{s}Y^a_{p s} \;. \lb{phiandchi}
\ee
Plugging this expansion into equations \eqref{ExcitationOperatorsDef3} and projecting on a arbitrary element of the basis provides a linear systems of equations for the coefficients
$X^a_{ps}$ and $Y^a_{ps}$, namely
\begin{align}
 \sum_{s'}(H_{0ss'}-\eps_p\delta_{ss'})X^a_{ps'} + \bra{s}\op V'[\op E_a]\ket{\psi_p} &= \Omega_a X^a_{ps} \;, \nn \\
 -\sum_{s'} (H_{0 s' s}-\eps_p\delta_{s s'})Y^a_{ps'} - \bra{\psi_p}\op V'[\op E_a]\ket{s}& = \Omega_a Y^a_{ps} \;. \nn
\end{align}
Using this basis, excitation operators are expressed as a linear combination of transition operators $\excite{p}{s} \equiv \ketbra{s}{\psi_p}$ and $\decay{s}{p} \equiv \ketbra{\psi_p}{s}$,
as follows
\be\lb{ExcitationOpBasisTransition1}
\op E_a = \sum_{ps}\excite{p}{s}X^a_{ps}+\decay{s}{p}Y^a_{ps} \;.
\ee
The contribution of the coupling operator in the equations of motion of excited states can be expressed by means of the coupling kernel \eqref{CouplingKernelDef1} as
\begin{align}
& \bra{s}\op V'[\op E_a]\ket{\psi_p} = \sum_{qs'}\left(U[\decay{s}{p};\excite{q}{s'}]X^a_{qs'} + U[\decay{s}{p};\decay{s'}{q}]Y^a_{qs'}\right) \;,\nn \\
& \bra{\psi_p}\op V'[\op E_a]\ket{s} = \sum_{qs'}\left(U[\excite{p}{s};\excite{q}{s'}]X^a_{qs'} + U[\excite{p}{s};\decay{s'}{q}]Y^a_{qs'}\right) \;, \nn
\end{align}
Given that we employ real functions both for $p$ and $s$ is real we also have $\excite{p}{s}(\r,\r') = \decay{s}{p}(\r',\r)$ and the density operator is symmetric in $\r \leftrightarrow \r'$.
These conditions applied together imply that $U[\excite{p}{s};\excite{q}{s'}] = U[\decay{s}{p};\decay{s'}{q}]$, as well as $U[\decay{s}{p};\excite{q}{s'}] = U[\excite{p}{s};\decay{s'}{q}]$,
and so equations \eqref{ExcitationOperatorsDef3} can be recasted in matrix form
\be\lb{ExcitationMatrixEq1}
\sum_{qs'}\mat{F_{ps}^{qs'} &  D_{ps}^{qs'}  \\
- D_{ps}^{qs'} & - F_{ps}^{qs'} }
\mat{X^a_{qs'} \\ Y^a_{qs'}} = \Omega_a \mat{X^a_{ps} \\ Y^a_{ps}}
\ee
where
\begin{align}
& F_{ps}^{qs'} = (H_{0 ss'}-\eps_p\delta_{ss'})\delta_{pq} + U[\decay{s}{p};\excite{q}{s'}] \;,\nn \\
& D_{ps}^{qs'} = U[\decay{s}{p};\decay{s'}{q}] \;.
\end{align}
This is the well-known Casida Equation in the basis of transitions for the excitation of energy $\Omega_a$. The computational reliability of the
eigenvalue $\Omega_a$ depends on the capability of the basis set $\ket{s}$ to fulfill of Eq.~\eqref{phiandchi}. Note that this may be valid only for a subset of the solutions of the eigenproblem \eqref{ExcitationMatrixEq1}.

\section{Computational details} \lb{compdetails}
The illustrative calculations presented in this paper are all KS-DFT calculations at the LDA level of theory. We employed the BigDFT code \cite{BigDFT},
that makes usage of Daubechies Wavelets as computational basis set. This orthonormal basis presents optimal features in view of reproducibility of the results, and it has proved to represent precisely and explicitly system with Free as well as Periodic boundary conditions (BC), without any basis set superposition error nor supercell aliasing, for Free BC.

For Ground State results, the important parameters in the context of convergence of the calculations are the spacing of the wavelet grid and the size of the simulation domain: for bound-state like functions, convergence is achieved by reducing the grid spacing and increasing the simulation box size.
Norm Conserving Pseudopotentials (PSP) of the HGH \cite{NLCCPSP} type are employed to remove the core electrons. Such PSP have proved to be able to provide all-electron accuracy for molecular systems.

@@

\subsection{Benzene}
The molecule is oriented in the $xy$ plain and the $x$-dimension of the box is equal to 15.0, 18.2 and 19.7 \AA.
220 states have been computed in each setup and the corresponding
energy range for the unoccupied states is 16.9, 11.6 and 9.7 eV.
